---
title: "Density of sil-co-sil-75 material"
date: "2022-03-31"
mainfont: Roboto
geometry: landscape
output:
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(magrittr)
library(kableExtra)

#options(knitr.kable.na = "")
options(knitr.kable.NA = '')
```


# Maximum density 

## Procedure 

-load material in a single lift 

- 15 drops from 12" height

- final reading should be between `r (3+1.57) - 0.2` and `r (3+1.57) + 0.2`
  - likely about `r round( pi * (2.54^2) * (3*2.54) * 1.45, -1)` g
- record uncorrected height reading (inches) and total mass (grams, including rubber band + cheesecloth)

- 5 reps of each; save material 


<!-- \clearpage -->

```{r max-data}

max_data_table <- list.files(path = here::here("ecmdata/raw-data/experiments-1-and-2/F1815-cylinders-data/sil-co-sil-75"), pattern = "max.*\\.csv$", recursive = T, full.names = T) %>%
  readr::read_csv(show_col_types = F, na = c("", "-")) %>%
  dplyr::select(-dplyr::matches("tin.*"), -c(date, ambient_temp_c, w_target))

max_data_table %>% 
  kbl(format = 'latex', caption = "Data collection for max density",
      longtable = TRUE) %>% 
  row_spec(row = 0, bold = TRUE) %>% 
  kable_styling(latex_options = c('hold_position', 'striped', 'repeat_header', 'scale_down'))
```


\clearpage 

# Minumum density 

## Procedure 

- gently fill cylinder using funnel

- flatten top w/ straight edge

- record total mass (including rubber band + cheesecloth)

```{r min-data}

min_data_table <- list.files(path = here::here("ecmdata/raw-data/experiments-1-and-2/F1815-cylinders-data/sil-co-sil-75"), pattern = "min.*\\.csv$", recursive = T, full.names = T) %>% 
  readr::read_csv(show_col_types = F, na = c("", "-"))  %>% 
  dplyr::select(-dplyr::matches("tin.*"), -c(date, ambient_temp_c, w_target))

min_data_table %>% 
  kbl(format = 'latex', caption = "Data collection for minimum density",
      longtable = TRUE) %>% 
  row_spec(row = 0, bold = TRUE) %>% 
  kable_styling(latex_options = c('hold_position', 'striped', 'repeat_header', 'scale_down'))
```
