---
title:  "Influence of coarse additions on the plasticity and toughness of soil mixtures, Part II: Effect of sand angularity and sand size uniformity."
date: "last compiled `r ecmfuns::ecm_timestamp()`"
mainfont: Roboto
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: false
    includes:
      in_header: preamble.tex
  officedown::rdocx_document:
    reference_docx: "oversizeAttLims-paper_word-doc-template.docx"
    number_sections: false
    tables:
      caption:
        pre: 'Table '
# bibliography: [library.bib, packages.bib]
# csl: soil-science-society-of-america-journal.csl
linestretch: 1.5
links-as-notes: false
papersize: letter
link-citations: true
urlcolor: blue
linkcolor: blue
toccolor: blue
mathfont: Fira Math Regular
---

<!-- Note that in the text of this paper, two experiments are referenced. -->
<!-- The original deal was having everything in one paper. So the experiments  -->
<!-- are numbered 1-3. All the data analyses was done with this in mind.  -->
<!-- So the dynamic number references in this paper all refer to experiments  2, and 3 -->
<!-- when in fact they are referring to what are (for this paper) termed  -->
<!-- experiments 1 and 2. A little confusing but leaving it this way  -->
<!-- will be much, much simpler than trying to re-do all the dynamic references, -->
<!-- figures, etc.  -->

\linenumbers



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F,
                    fig.align = 'center',
                    dpi = 72,
                    out.width = '90%')

knitr::opts_knit$set(root.dir = here::here())

options(
  knitr.graphics.rel_path = FALSE,
  knitr.kable.NA = "-"
)

library(purrr)
```


<!-- need to load some variables that will be dynamically referenced in the abstract  -->


```{r load-uniformity-experiment-Cu-values}
# read and unload the list into global environment, then delete the 
# original list to reduce environment clutter
uniformity_experiment_d_x_values <- readr::read_rds(
file = here::here(
  "ecmdata/derived-data/dynamic-number-references/uniformity-experiment-dx-values.rds"
)
)

uniformity_experiment_d_50_values <- uniformity_experiment_d_x_values$uniformity_experiment_d_50_values
uniformity_experiment_Cu_values <- uniformity_experiment_d_x_values$uniformity_experiment_Cu_values
```

```{r load-max-testable-sand-contents}
max_testable_sand_contents <- readr::read_rds(
  file = "./ecmdata/derived-data/dynamic-number-references/uniformity-experiment-max-testable-sand-contents.rds") %>% 
   map(~format(100 * ., digits = 3, drop0trailing = TRUE))

```


```{r load-shape-experiment-pure-clay-atterberg-limit-values}

shape_experiment_pure_clay_atterberg_limit_values <- readr::read_rds(
here::here('ecmdata/derived-data/dynamic-number-references/shape-experiment-pure-clay-atterberg-limit-values.rds'))

```


```{r load-uniformity-experiment-pure-clay-atterberg-limit-values}
uniformity_experiment_pure_clay_atterberg_limit_values <- readr::read_rds(
here::here('ecmdata/derived-data/dynamic-number-references/uniformity-experiment-pure-clay-atterberg-limit-values.rds')
)
```


# Abstract 

The Atterberg limits may offer a useful means to evaluate baseball infield soils because they quantitatively relate soil behavior to water content. 
Prior research has demonstrated that liquid and plastic limits (LL and PL) of sand-clay mixtures are affected by the quantity and type of admixed sand, but these studies have used sand exclusively <425 &mu;m and little attention has been devoted to sand angularity and sand-size uniformity. 


This research was conducted to clarify the effect of sand angularity and sand-size uniformity on the Atterberg limits of soil mixtures containing a range of sand contents and a significant mass percentage 425-2000 &mu;m. 


Experiment 1 compared the effect of mixing either an angular or a round sand (both 0.5-1 mm) with a kaolinitic clay at sand contents between 0 and 80%. 
Little difference was observed in LL and PL, suggesting angularity plays a minimal role on mix performance.


<!-- Experiment 2 compared the effect of sand-size uniformity on LL and PL.  -->
Experiment 2 compared the effect of mixing one of two sands having similar D~50~ (`r uniformity_experiment_d_50_values[["low_Cu"]]` and `r uniformity_experiment_d_50_values[["high_Cu"]]` mm) but varying uniformity (uniformity coefficients of `r uniformity_experiment_Cu_values[["low_Cu"]]` vs. `r uniformity_experiment_Cu_values[["high_Cu"]]`) with an illitic clay at sand content 0-80%.
<!-- The high-C~u~ sand had slightly higher LL and PL at equivalent sand content. -->
Mixtures including the high-C~u~ sand maintained their plasticity to higher sand content (~`r max_testable_sand_contents$high_Cu_max_rollable_sand_pct`%) than those produced with the low-C~u~ sand (~`r max_testable_sand_contents$low_Cu_max_rollable_sand_pct`%).

Calculations for the threshold fines content (TFC) and intergranular porosity agreed closely with the experiments, indicating a potential to estimate TFC from sand porosity alone. 


<!-- This research demonstrated the effect of sand angularity on LL and PL was minimal. -->
<!-- The uniformity effect was measurable, but it was small compared with that of total sand content. -->
<!-- Uniformity had a more notable influence on the sand content separating plastic and nonplastic mixtures.  -->



# Introduction 


The Atterberg limits are used by geotechnical engineers to classify fine-grained soils [@ASTMInternational2017]. 
These tests include the liquid limit (LL) and plastic limit (PL).
The tests were developed for fine-grained soils, but research has shown the amount and properties of admixed sand also affects the test results [@Dumbleton1966b; @Sivapullaiah1985; @Barnes2013; @Atterberg1911]. 
Particle angularity and particle-size uniformity are known to affect the behavior of granular materials, but their effects are less clear for soils which contain significant amounts of fines  [@Holtz2010; @Miller2011; @Mitchell1993; @Hojae2001; @Zuo2015].


<!-- ## Prior research on Atterberg limits of soils having varied sand angularity   -->
## Atterberg limits of soils having varied sand angularity  

Particle angularity is defined as the roughness of a particle surface [@Brady2007]. 
<!-- Mathematical definitions are used in a variety of scientific areas,  -->
Mathematical definitions have been used to quantify angularity, although it is often estimated using representative charts developed from the metrics  [@Wadell1932; @Suhr2020; @Krumbein1941].


Research on the importance of sand angularity in soil mixtures has produced mixed results. 
@Dumbleton1966b tested mixtures of various types of coarse particles with either kaolinite or montmorillonite. 
@Dumbleton1966b demonstrated that mixtures including angular sand had higher LL and PL relative to mixtures produced with equivalent amounts of glass spheres.
The researchers postulated that angular particles had more surface area compared to the spheres, requiring additional water to coat their surfaces before they could slide or flow past one another.


Findings by @Sivapullaiah1985 differ from those of @Dumbleton1966b.
@Sivapullaiah1985 mixed either angular or round particles of the same size fraction (150 μm – 75 μm) with bentonite clay at sand contents ranging from 20-95 %. 
@Sivapullaiah1985 reported no difference in the liquid or plastic limits due to angularity.

## Atterberg limits of soils having varied sand uniformity

Particle-size uniformity, or more simply "uniformity," is defined as the similarity of diameters across particles within a soil. 
The coefficient of uniformity (C~u~) is a commonly used mathematical definition [@Holtz2010; @Adams1994]. 
C~u~ is the ratio between the particle diameters at which 60% and 10% of the sample is finer:

\begin{equation}
C_u = \frac{D_{60}}{D_{10}}
(\#eq:Cu-equation)
\end{equation}

Little research is available on the effect of sand uniformity on the Atterberg limits of soil mixtures.
This is probably because the upper boundary of allowed particle sizes in ASTM D4318 (425 &mu;m) limits the total range of sand particle diameters.
@Dumbleton1966b reported that mixes produced using non-uniform sand had similar Atterberg limits to mixes containing sand from a single mesh size.
However, all of the sand in their experiment was between 425 and 53 &mu;m, limiting the potential variability in uniformity. 
Efficient particle packing - defined as the ability of smaller particles to fit comfortably in the voids between larger particles - is limited when the majority of particle diameters fall within a factor of 10 [@Lade1998].
It is possible that a larger uniformity effect on LL and PL would be observed if coarser sand were included in the mixture to create a higher C~u~.

<!-- Sand-size uniformity may play a role in governing the behavior of mixed-grain soils.  -->
At very low sand contents, angularity and uniformity are probably unimportant because the sand grains are suspended in a clay-water matrix and do play an insignificant role in transmitting loads through the soil skeleton. 
At higher sand contents, the coarse grains begin to contact one another and the nature of these contacts will affect the behavior of the bulk soil [@Zuo2015; @Mitchell1993].
As sand content continues to increase, the soil becomes nonplastic because plasticity is not observed in soils dominated by granular particles [@Holtz2010; @Mitchell1993].
The behavior transition between sand-like and fines-like behavior occurs at a fines content termed the threshold fines content (TFC) [@Zuo2015; @Sibley2020].


Because less-uniform sands have less total void space, one could expect these sands to require a lesser proportion of fines to fill their interstices, compared with a more uniform sand. 
Therefore, sand uniformity could alter the upper limit of sand content at which the soil will exhibit plasticity.


<!-- The authors are unaware of any other research on mixtures containing sand with lower C~u~.  -->

# Objectives 

<!-- The Atterberg limits may offer a useful way to evaluate baseball infield soils because soil water content is an important consideration for field managers [@Schroder2009; @Zwaska1999]. -->
Recent research has demonstrated that the LL and PL tests can be performed on soils containing a significant amount of particles 2000-425 &mu;m  (Mascitti and McNitt, this issue).
<!-- Baseball infield soils often contain significant quantities of sand in this range. -->
It is unclear what effect sand shape and uniformity play on the Atterberg limits of soil mixtures containing these coarser sands.
<!-- , and how they may translate to the the behavior of infield soils in situ. -->

The present research was conducted to answer two questions:

1. How does sand particle angularity affect LL and PL of sand-clay mixtures when total sand content and sand particle size are held constant?


1. How does sand-size uniformity affect LL and PL of sand-clay mixtures when total sand content and _average_ sand particle size (D~50~) are held constant?



<!-- 1. How does sand particle angularity (angular vs. round) affect LL and PL of sand-clay mixtures when total sand content and sand particle size are held constant? -->


<!-- 1. How does sand-size uniformity (uniform vs. non-uniform) affect LL and PL of sand-clay mixtures when total sand content and _average_ sand particle size (D~50~) are held constant? -->


<!-- 1. Do the effects of particle angularity or uniformity interact with sand content? -->


<!-- 1. Do the effects of particle angularity or uniformity change as a function of sand content? -->


# Materials and methods

Two experiments were conducted to evaluate the effects of sand angularity and uniformity on the Atterberg limits of soil mixtures containing a range of sand contents.


## Mixture component characterization

The minimum and maximum void ratios of the sand components used in each experiment were characterized using modified versions of ASTM [-@ASTMF1815-11; -@ASTMInternational2016].
In the minimum void ratio test , a single lift of sand was dynamically compacted using a specified drop weight. 
In the maximum void ratio test, the sand was carefully poured into a container of known volume and leveled with a straight edge. 


The liquid and plastic limit of each clay component were determined individually using ASTM D4318 [@ASTMD43182018].
The LL of the kaolinitic clay used in Experiment 1 was `r shape_experiment_pure_clay_atterberg_limit_values$LL` and its PL was `r shape_experiment_pure_clay_atterberg_limit_values$PL`, yielding a plasticity index of `r shape_experiment_pure_clay_atterberg_limit_values$PI`.
The PL of the illitic clay soil used in Experiment 2 was `r uniformity_experiment_pure_clay_atterberg_limit_values$PL` and its LL was `r uniformity_experiment_pure_clay_atterberg_limit_values$LL`, yielding a PI of `r uniformity_experiment_pure_clay_atterberg_limit_values$PI`. 
The particle size distribution of the illitic clay soil used in Experiment 2 is also shown in Figure \@ref(fig:uniformity-experiment-particle-size-curves).




## Mixing procedure

All mixture components were air-dried and their water contents were determined gravimetrically.
The clay components were pulverized and passed though 0.25 mm screen.
Sands were mixed by hand with the relevant clay component until visually homogeneous.
The mixture component percentages were adjusted for the trace amounts of particles 2000-53 &mu;m in the clay components. 
Final mixture percentages are expressed as oven-dry mass.


## Treatments 

Experiment 1 evaluated mixtures containing equivalent amounts of one of two sands having varying angularity.
Two sands were selected based on their classification as angular and well-rounded.
The sands were visually classified using a microscope and the chart from @Baker2006a (Figure \@ref(fig:sand-photomicrographs)). 
Each sand was then repeatedly sieved to remove particles <0.5 mm and >1 mm. 
The single-mesh sands were mixed with a kaolinitic clay to yield mixtures having 0, 20, 40, 60, 70, 75, and 80 % sand.


```{r sand-photomicrographs, fig.cap='Angular (A.) and well-rounded (B.) sands used for Experiment 1.', eval=F}
knitr::include_graphics(
  here::here(
'figs', 'png', 'image-figs', 'sand-photo-micrographs.png')
)
```

```{r load-uniformity-experiment-dx-values}

uniformity_experiment_d_x_values <- readr::read_rds(
file = here::here(
  "ecmdata/derived-data/dynamic-number-references/uniformity-experiment-dx-values.rds"
)
)

uniformity_experiment_d_50_values <- uniformity_experiment_d_x_values$uniformity_experiment_d_50_values
uniformity_experiment_Cu_values <- uniformity_experiment_d_x_values$uniformity_experiment_Cu_values

```


Experiment 2 evaluated the effect of sand-size uniformity for mixtures containing one of two sands having similar D~50~ but varying C~u~.
The two sands were produced from a single washed concrete sand having a wide particle size distribution. 
The original concrete sand was sieved to remove all particles >1 mm and <0.25 mm.
The remaining fraction between 1 and 0.5 mm was riffled into two aliquots.
The first aliquot remained untouched and is termed the "low-C~u~ sand".
The second aliquot, termed the "high-C~u~ sand", had a portion of the removed particles (>1 mm and <0.25 mm) returned.
This procedure created a wider particle-size distribution while maintaining a similar D~50~.
Figure \@ref(fig:uniformity-experiment-particle-size-curves) shows the particle size distributions of the low-C~u~ and high-C~u~ sands.
These sands meet the criteria of having similar D~50~ values (`r uniformity_experiment_d_50_values[["low_Cu"]]` mm and `r uniformity_experiment_d_50_values[["high_Cu"]]` mm) but different C~u~ values (`r uniformity_experiment_Cu_values[["low_Cu"]]` vs. `r uniformity_experiment_Cu_values[["high_Cu"]]`). 
While the C~u~ value of `r uniformity_experiment_Cu_values[["high_Cu"]]` is still relatively low compared to natural alluvial sands, in this experiment the maximum obtainable C~u~ was limited by the maximum particle diameter of 2000 &mu;m and the intentionally limited mass of particles <425 &mu;m. 



<!-- Treatments included sand content and C~u~ and each sand was mixed several ratios ... -->


Each of these sands were mixed with a single illitic clay soil to yield mixtures having between 0 and 80% sand, for a total of 42 mixtures. 
Mixtures were produced at 5% sand content intervals between 0 and 50%, and a 2.5% interval between 50 and 80% sand.
The increased spacing of data points between 50-80% sand was designed to provide better resolution near the threshold fines content. 




```{r uniformity-experiment-particle-size-curves, fig.cap='Particle size distributions for the two sands and one clay used in Experiment 2. Dashed grey lines indicate $D_{50}$ for each sand.', eval=F}
ecmfuns::insert_fig('uniformity-experiment-particle-size-curves')
```


## Atterberg limit test protocol

After mixing the sand with the relevant clay component, LL and PL tests were performed on a series of mixtures to evaluate the effect of sand angularity or uniformity.

The liquid and plastic limit tests were performed according to a modified version of ASTM D4318 [@ASTMD43182018].
The modification eliminated the wet-sieving procedure so particles between 2000 and 425 &mu;m remained in the sample (Mascitti and McNitt, this issue). 
At least four data points were collected during the LL test in order to plot the flow curve.
In the PL test, 3 threads were rolled to the crumbling condition before being weighed to &pm; 0.001 g and oven-dried. 
In both experiments the average of the 3 results was used to represent the PL of each sample.


## Statistical analysis and computational environment

LL and PL were the dependent variables in both experiments. 
A two-way ANOVA model was fitted to test the interaction effect between particle shape and percent sand (Table \@ref(tab:shape-experiment-anova-table))
In Experiment 1, particle shape was considered a categorical predictor while a 2nd-order polynomial spline term was used to model percent sand as a continuous predictor. 
<!-- consuming a single degree of freedom (i.e. covariate). -->
In Experiment 2, C~u~ was considered a categorical predictor while percent sand was treated as in Experiment 1. 
Main effects and interactions were tested using Type III Sums of Squares.
Treatments were considered significantly different at &alpha; = 0.05.


All analyses were performed using the `lm()` function in the R Language for Statistical Computing (version `r stringr::str_extract(sessionInfo()$R.version$version.string, "(?<=R\\sversion\\s)[\\dd\\.]*(?=\\s)")`) [@R-Core-Team2022].
GNU `Make` [@GNU2020] was used to facilitate reproducible analyses by maintaining links between raw data, analysis code, and finished output.
Raw data and analysis code are included in the supplemental materials. 

<!-- \clearpage -->


# Results and discussion


## Experiment 1: Effect of sand angularity

Experiment 1 compared the effect of angularity for two sands of equal size ranging from 0-80% sand. 
Figure \@ref(fig:shape-experiment-atterberg-limit-facets) shows that the LL and PL were nearly identical for sand content <60%.
At sand content &ge; 60%, a very slight increase in both LL and PL is visible for the angular sand, but this increase was not statistically significant.


```{r experiment-2-dynamic-numbers}
largest_experiment_2_diff <- readr::read_rds(
here::here(
  'ecmdata/derived-data/dynamic-number-references/experiment-2-differences.rds'
)
)
```


<!-- The interaction effect between particle shape and sand percent was small and not statistically significant. -->
The maximum difference between angular and round sand for any of the tests was `r largest_experiment_2_diff` % water content.
Under the conditions of this study, particle angularity is not important when sand size and sand contents are held constant. 


```{r shape-experiment-atterberg-limit-facets, fig.cap='The effect of particle shape on both the LL and PL tests was minimal.', eval=F}
ecmfuns::insert_fig('experiment-2-atterberg-limit-facets')
```


```{r shape-experiment-anova-table, eval=FALSE, tab.cap="Analysis of variance table for each characteristic water content in Experiment 1.", tab.id ="shape-experiment-anova-table", tab.cap.style = "Table Caption"}

# This effect was not significant and I don't think
# it's important to show, as the difference is 
# obviously small. Leaving the code here in case omebody asks me to put the table in. 

# update, May 5th, I put it back in now that we have split the paper up into 2 parts.

source(
here::here(
  'src/R/table-generation/shape-experiment-ANOVA-table.R'
)
)
if(knitr::is_latex_output()){shape_experiment_anova_tables$latex_kable}
if(knitr::is_html_output()){shape_experiment_anova_tables$html_kable}
if(!knitr::is_html_output() && !knitr::is_latex_output()){shape_experiment_anova_tables$word_flextable}
```

## Experiment 2: Effect of sand-size uniformity


### Uniformity effect on LL and PL 


```{r uniformity-experiment-sand-contents-count}

# as of 2022-05-10 not using this any more...not really important to count 
# them as the point is pretty obvious from the plots.

uniformity_experiment_atterberg_limits_data <- readr::read_rds(here::here('ecmdata/derived-data/cleaned-rds-files/uniformity-experiment-atterberg-limits-cleaned-data.rds') )

n_uniformity_experiment_sand_contents <- uniformity_experiment_atterberg_limits_data %>% 
purrr::pluck('sand_pct') %>% 
unique() %>% 
length()

n_uniformity_experiment_high_Cu_w_exceeding_low_Cu <- uniformity_experiment_atterberg_limits_data %>% 
tidyr::pivot_wider(
  id_cols = c(sand_pct, test_type),
  names_from = sand_name,
  values_from = water_content
)%>% 
dplyr::mutate(
  high_Cu_w_greater = high_Cu_Quickcrete > low_Cu_Quickcrete
) %>% 
dplyr::group_by(test_type) %>% 
dplyr::filter(high_Cu_w_greater) %>% 
dplyr::count()
n_uniformity_experiment_high_Cu_w_exceeding_low_Cu_LL <- purrr::pluck(n_uniformity_experiment_high_Cu_w_exceeding_low_Cu[n_uniformity_experiment_high_Cu_w_exceeding_low_Cu$test_type == "LL", 'n'], 'n')
n_uniformity_experiment_high_Cu_w_exceeding_low_Cu_PL <- purrr::pluck(n_uniformity_experiment_high_Cu_w_exceeding_low_Cu[n_uniformity_experiment_high_Cu_w_exceeding_low_Cu$test_type == "PL", 'n'], 'n')

```


<!-- The LL and PL of both mixtures including either sand were inversely proportional, up tpo to the higher end of the range of sand contents tested. -->
The LL and PL of both mixtures including either sand were inversely proportional to sand content, up to the higher end of the range (~65-70%%), above which the relationship is less clear. 
Both sands in Experiment 2 were relatively coarse (D~50~ of  `r uniformity_experiment_d_50_values[["low_Cu"]]` and `r uniformity_experiment_d_50_values[["high_Cu"]]` mm), and the strong inverse relationship agrees with other research on mixtures containing coarse sand (Mascitti and McNitt, Part I, this issue).

Mixtures produced with low-C~u~ sand had higher LL and PL than those produced using high-C~u~ sand (Figure \@ref(fig:uniformity-experiment-atterberg-limit-facets)).
This small effect was statistically significant for the LL but not for the PL (Table \@ref(tab:uniformity-experiment-anova-table)).
Although this effect is measurable, it is probably of little practical significance.
In Experiment 2, the sand content played a more dominant role on LL and PL than sand uniformity.
While only a single clay was tested, the effect of varying the clay's plasticity would likely affect LL and PL to a greater degree than varying the C~u~ of the sand.

@Dumbleton1966b suggested that observed differences in LL and PL due to sand particle size were attributable to higher specific surface area (SSA).
The nonlinear relationship between SSA and particle diameter dictates means requires that the low-C~u~ sand would consume more water and increase the water content of the clay matrix.


```{r uniformity-experiment-atterberg-limit-facets, fig.cap='The high-Cu sand had higher LL and PL than the low-Cu sand at nearly all sand contents.', eval=F}
ecmfuns::insert_fig('uniformity-experiment-atterberg-limit-facets')
```

```{r uniformity-experiment-anova-table, tab.cap="Analysis of variance table for each characteristic water content in Experiment 2. Significant effects at α=0.05 in bold.", tab.id ="uniformity-experiment-anova-table", tab.cap.style = "Table Caption", eval=FALSE}
source(
here::here(
  'src/R/table-generation/uniformity-experiment-ANOVA-table.R'
)
)
if(knitr::is_latex_output()){uniformity_experiment_tables$latex_kable}
if(knitr::is_html_output()){
  uniformity_experiment_tables$html_kable
  } else{uniformity_experiment_tables$word_flextable}
```


### Uniformity effect on threshold fines content

A more interesting feature of Experiment 2 was that the mixtures containing different C~u~ sands became nonplastic at different sand contents.

Low-C~u~ sand mixes containing 70, 72.5, and 75% sand could not be rolled into threads, but high-C~u~ mixes at these sand contents still had measurable plastic limits.
<!-- Similarly, at sand content 75-80 %, the low-C~u~ mixtures slid along the base of the Casagrande cup (invalidating the test), whereas high-C~u~ sand mixtures flowed (rather than slid) and therefore had measurable liquid limits. -->
These observations suggest that the threshold fines content is affected by sand uniformity and merits further study.


<!-- In its pure state, non-uniform sand has less void space because the smaller sand particles fill some of the interstices among larger grains.  -->
<!-- This inter-packing reduces the volume of fines and water required to fill the voids, allowing the sand particles to begin losing intergranular contact at a higher total sand content. -->


Figure \@ref(fig:tfc-by-sand-uniformity-drawing) shows conceptual diagrams of the particle arrangement at the TFC, along with phase diagrams depicting the volume occupied by each of the four phases (sand solids, clay solids, water, and air).
When interpreting Figure \@ref(fig:tfc-by-sand-uniformity-drawing) it is important to recall that sand content is normalized to mass while void ratio is normalized to volume. 
Total bulk volume is equivalent (i.e. 100%), but the mass fraction of sand is larger for the non-uniform sand when the sand content is normalized to the total solids mass (i.e. $\frac{m_{sand}}{m_{sand} + m_{clay}}$).
<!-- Mass cannot be drawn and that the diagrams represent volumes normalized to 1 (i.e. 100%).
-->


Calculations for the TFC and real soil behavior observed in experiments rarely coincide [@Zuo2015]. 
However, little attention has been paid to soil mixtures containing plastic fines as opposed to nonplastic silt or two granular materials.
While not replicated or tested across multiple clay types, this research has demonstrated the potential to accurately compute TFC from the minimum void ratio of a sand without the need to perform time-consuming experiments. 



```{r tfc-by-sand-uniformity-drawing, fig.cap='Mixtures including less-uniform sand (left) should theoretically require a lesser volume of fines + water to fill their voids than mixtures containing more uniform sand (right).', eval=FALSE} 
if(knitr::is_latex_output()){knitr::include_graphics(
  path = here::here("illustrations", "tfc-by-sand-uniformity", "uniformity-effect-on-transitional-fines-content.pdf")
)}else{
 knitr::include_graphics(
  path = here::here("illustrations", "tfc-by-sand-uniformity", "uniformity-effect-on-transitional-fines-content.png")
) 
}
```



To test the concept shown in Figure \@ref(fig:tfc-by-sand-uniformity-drawing), maximum void ratios $e_{sa~(max)}$ for each sand were used to compute the sand content corresponding to the TFC (i.e. $m_{sa} = 1 - m_{fines}$).
This state theoretically represents the sand content at which the sand grains begin to lose contact in the soil skeleton [@Mitchell1993].
The sand content at this boundary is computed as:

\begin{equation}
m_{sa} = \frac{1 +  \frac{\frac{w G_c }{G_w}}{S_e}}{1 + \frac{e_{sa~(max)} G_c}{G_{sa}}}
(\#eq:tfc-equation)
\end{equation}

where $m_{sa}$ is the sand mass, $w$ is the gravimetric water content, $S_e$  is the effective saturation, $G_w$ is the specific gravity of water at 20 &deg;C, $G_{sa} and G_c$ are the specific gravities of the sand and clay, and $e_{sa}$ is the void ratio of the sand. 
A full derivation of Equation \@ref(eq:tfc-equation) is available in the supplemental materials.

Effective saturation $S_e$ is the percent of voids occupied by water, normalized to the total porosity [@Holtz2010].
$S_e$ must be accounted for in Equation \@ref(eq:tfc-equation) because it is unrealistic to expect complete air evacuation by simply compressing the soil with one's fingers [@Haigh2013]. 
No attempt was made to measure $S_e$ directly from soil threads.
The value of 0.9 was chosen based on literature regarding soil compaction tests; compacting a soil slightly wet of optimum typically yields $S_e$~0.9, regardless of compaction effort [@Holtz2010; @Gurtug2015; @Rahman2022].
There is also a good relationship between PL and the Proctor optimum water content, with PL being ~1.1$w_{opt}$ [add refs here , Barnes is one but there are others].
Therefore, the soils in this experiment would qualify as wet-of-optimum and would likely have  $S_e$~0.9.


```{r load-tfcs-for-each-sand}

computed_tfcs <- readr::read_rds(
  file = here::here(
    "ecmdata", "derived-data", "dynamic-number-references", 
    "uniformity-experiment-TFC-calculations-from-void-ratio.rds")
) %>% 
  map(~as.character(round(100 * ., digits = 0)))

```


This calculation yielded sand contents of `r computed_tfcs$low_Cu_tfc`% for the low-C~u~ sand and `r computed_tfcs$high_Cu_tfc`% for the high-C~u~ sand. 
Recall that mixtures containing these sands became nonplastic at sand content of `r max_testable_sand_contents$low_Cu_max_rollable_sand_pct` and `r max_testable_sand_contents$high_Cu_max_rollable_sand_pct` respectively. 
Given the elementary nature of the PL test, the agreement between the calculations and experimental results is rather striking.
<!-- Given the elementary nature of the PL test, the close agreement between the calculations and experimental results is rather striking. -->
This finding suggests TFC may be accurately computed from $e_{sa~(max)}$.


The same calculation was made at the measured LL for the mixes containing the highest sand content at which the LL test could be performed without the soil sliding in the cup.
Research has shown that even at the LL, $S_e$ is not complete and that $S_e$ asymptotically approaches 1 with increasing water content [@Guo2022; @Holtz2010].
Therefore $S_e$=0.95 was used along with LL water contents and Equation \@ref(eq:tfc-equation) to yield calculated TFC values of `r computed_tfcs$high_Cu_max_LL_testable_sand_pct` and `r computed_tfcs$high_Cu_max_LL_testable_sand_pct` for the low-C~u~ and high-C~u~ sands.
These closely match the sand contents at which the LL test became invalid (`r `) and `r `


<!-- Given the sometimes perceived crudeness of the Atterberg limit tests, the agreement between the calculated and experimental results is rather striking.  -->


A final means to evaluate the status of the sand particles is intergranular porosity $n_{sa}$.
This parameter represents the void volume between the sand grains if the fines are treated as voids along with the water and air [@Lade1998a].
It can be assumed that if $n_sa$ exceeds the porosity of the sand alone, the sand grains have begun to separate into a less-stable configuration. 
The $n_{sa} can be computed as:

\begin{equation}
n_{sa}~=~\frac{\frac{m_{c}}{G_{c}}~~+~\left[\frac{w}{G_w}~+~\frac{\frac{w}{G_w}~\cdot~\left({1-S_e}\right)}{S_e}\right]}{\frac{m_{sa}}{G_{sa}}~+~\frac{m_{c}}{G_{c}}~+~\left[\frac{w}{G_w}~+~\frac{\frac{w}{G_w}~\cdot~\left({1-S_e}\right)}{S_e}\right]}
(\#eq:sand-phase-porosity-equation)
\end{equation}

where $m_c$ is the mass of the clay fraction, $m_{sa}$ is the mass of the sand fraction, $w$ is the water content, $S_e$ is effective saturation, and $G_sa$, $G_c$, and $G_w$ are the specific gravities of the sand, clay, and water phases. 

Values for $n_sa$ were computed for the observed plastic limit water contents.
Again, $S_e$ was assumed to be 0.9.
Figure \@ref(fig:sand-void-ratio-plot) plots $n_{sa}$ as a function of sand content for mixtures including either type of sand.
The horizontal lines represents the maximum porosity of each sand with no fines added. 
Again, the computed $n_{sa}$ values agree closely with those observed at the sand contents required to make the soil nonplastic. 


```{r sand-void-ratio-plot, fig.cap='Computed values for intergranular agreed closely with the sand content required to make the mixture nonplastic.', eval=FALSE}
ecmfuns::insert_fig('uniformity-experiment-sand-void-ratio-vs-sand-content')
```



<!-- Do I need to compute SSA for the two sands?? I can do it if I first fit the PSD curves, derive the parameters assuming lognormality, and then convert particle diameter to SSA and take the integral of the curve.  -->

<!-- However, this seems like a lot of work for a pretty small conclusion... -->
<!-- but maybe not.  -->

<!-- A uniform sand has greater porosity in its densest state.  -->
<!-- For equivalent mass of fines, this leaves more room for water to fill the voids and allow the particles to slide past one another, whether it causes slumping failure at at the liquid limit or brittle failure at the plastic limit.  -->




# Conclusions


<!-- In this research, the effect of particle angularity on Atterberg limits of sand-clay mixtures was negligible and the effect of sand-size uniformity was statistically significant but of little practical importance.  -->
<!-- small butstatistically significant. -->

<!-- This research demonstrated that the effect of particle angularity on Atterberg limits of sand-clay mixtures is negligible and that the effect of sand-size uniformity is small, but more notable. -->

In Experiment 1, the LL and PL of mixtures containing equivalent sand content but varying unifority were virtually identical. 


In Experiment 2, Low-C~u~ mixtures had significantly higher LL compared to high-C~u~ mixtures having equivalent sand content. 
There was an analogous trend for PL, but the effect was not significant. 
The higher LL and PL could be due to the higher specific surface area of the high-C~u~ sand. 


Mixtures produced with the low-C~u~ sand became nonplastic at relatively higher sand content.
This suggests the threshold fines content (marking a transition between plastic and nonplastic behavior) could potentially be computed from the void ratio of the sand.


<!-- Less-uniform sands have fewer voids because smaller sand grains can fill the interstices between larger grains.  -->
<!-- This reduces the void volume available for the fines, which allow the granular particles to slide past one another and permit the soil to exhibit plasticity.  -->
<!-- Calculations for the threshold fines content and intergranular porosity agreed closely with observed LL and PL values.  -->
<!-- This suggests the physical properties of the sand may be used to accurately accurately calculate the fines content marking a transition between plastic and nonplastic behavior.  -->


Future work should investigate the potential for relating the threshold fines content to Atterberg limits, and predicting TFC from a sand's maximum void ratio. 
Additional research could evaluate potential interactions between sand uniformity and the plasticity of the clay, and investigate mixtures with higher C~u~ sands. 

<!-- Higher plasticity clays may consume more water and  -->

<!-- The results of this research can aid soil suppliers and grounds managers in measuring and predicting the performance of baseball infield soils.  -->


# Conflict of interest 

The authors declare that there are no conflicts of interest. 

# Author contributions 

<!-- use the tenzing package to auro-generate the Markdown text  -->
<!-- see <https://marton-balazs-kovacs.github.io/tenzing/articles/local_use.html> -->
```{r author-contributions}

#tenzing::read_contributors_table()
#contributions_tbl <- 

#print_credit_roles(contributors_table = my_contributors_table, initials = TRUE, order_by = "contributor")
```



# References
